---
title: "R for Data Science - Class 1"
author: "Eric KWIZERA"
date: "2025-09-16"
output: html_document
---

# Important Definitions in Data Science

- **Statistics**
- **Modelling**
- **Computation**
- **Calculus**

# Today’s Class Topics
- Installation of R
- Installation of RStudio
- Calculator in R
- Data Importation in R
- Data Description (Numerical Measures and Graphs)
- Data Structures (Vectors, Factors, Lists, Data Frames, Matrices, Arrays)

# Vector: The Foundational R Data Structure

## Numeric
```{r}
x <- c(0.5, 0.6)
age <- c(20,35,32,29)
summary(age)
age <- c(35,24,18,24)
mean(age)
```

## Logical
```{r}
x <- c(TRUE, FALSE)
x <- c(T, F)
```

## Character
```{r}
x <- c("a", "b", "c")
class <- c("M", "F", "F", "M")
```

## Integer
```{r}
x <- 9:29
```

## Complex
```{r}
x <- c(1+0i, 2+4i)
```

# Matrices and Operations
```{r}
x <- matrix(1:6, nrow = 2, ncol = 3)
x

y <- matrix(1:6, nrow = 3, ncol = 2)
x %*% y
```

# Lists
```{r}
x <- list(1, "a", TRUE, 1 + 4i)
x
```

# Factors
```{r}
x <- factor(c("yes", "yes", "no", "yes", "no"))
```

# Handling Missing Data
```{r}
x <- c(1, 2, NA, 10, 3)
is.na(x)
```

# Coercion
```{r}
sum(c(TRUE, TRUE, FALSE, FALSE, FALSE))
```

# Creating Data Frames
```{r}
studentID <- c(1,2,3,4,5)
math_score <- c(12,17,10,9,NA)
gender <- c("M","F","M","M","F")
it_score <- c(13,18,11,10,19)
scoredata <- data.frame(studentID, gender, math_score, it_score)
scoredata
```

# Data Importation
```{r}
data_class <- read.table("E:/Datasets/New folder/Book1.txt")
variable.names(data_class)
head(data_class)
tail(data_class)
summary(data_class$WEIGHT)
```

# Using Built-in Datasets
```{r}
data("airquality")
variable.names(airquality)
str(airquality)
```

# Exploratory Data Analysis
```{r}
summary(data_class)
apply(data_class, 2, mean)
apply(data_class, 2, sd)
```

# Relationships and Visualization
```{r}
data("iris")
by(iris$Petal.Length, iris$Species, mean)
boxplot(iris$Sepal.Length ~ iris$Species, data = iris, main = "Comparison")
```

# Relationship Between Two Categorical Variables
```{r}
ucba <- data.frame(UCBAdmissions)
cross <- xtabs(Freq ~ Gender + Admit, data = ucba)
prop.table(cross, 2)
```

# Bar Chart
```{r}
dat <- read.table(text ="ProdA ProdB ProdC ProdD
1 110 50 60 70
2 120 50 80 65", header= TRUE)
barplot(as.matrix(dat), beside = FALSE, col = c("Red", "Green"))
```

# Scatter Plot
```{r}
plot(WEIGHT ~ HEIGHT, data = data_class)
cor(data_class$HEIGHT, data_class$WEIGHT)
```

# Linear Models
```{r}
data_lm <- lm(WEIGHT ~ HEIGHT, data = data_class)
summary(data_lm)
hist(data_class$HEIGHT)
hist(data_class$WEIGHT, col = "blue")
```

# Data Management: Create New Variable
```{r}
attach(data_class)
BMI <- round(WEIGHT / (HEIGHT / 100)^2, 2)
cbind(data_class, BMI)
summary(BMI)
stem(BMI, scale = 2)
```

